The Ranked Nearest Neighbour Interchange space

### 2019-08-14 paper.tex
edits on section: caterpillar trees

\begin{lemma}
    $\Delta(\rnni) \leq \frac{(n-1)(n-2)}{2}$, where $\Delta(G) = \max\limits_{u,v \in V(G)}d(u,v)$ denotes the diameter of graph $G$.
\end{lemma}

@@

In this section we restrict our attention to the set of \emph{caterpillar trees} where each internal node has at least one child that is a leaf.
This type of tree is of particular interest in $\rnni$ because shortest paths between such trees in $\rnni$ differ from those in classical $\nni$ space.
This in combination with the importance of caterpillar trees in the proof of $\np$-hardness of $\nni$ distances in \autocite{Dasgupta2000-xa} suggests that understanding the geometry of paths between caterpillar trees helps investigating the complexity of computing $\rnni$ distances.
In particular, we are going to prove that one can always find a shortest path between two caterpillar trees that contains caterpillar trees only, meaning that the set of caterpillar trees is convex in $\rnni$.
We discovered this by computing caterpillar paths using an algorithm to compute shortest paths within the set of caterpillar trees and comparing the results with shortest path in $\rnni$.
This algorithm will be explained in detail later in this section, after demonstrating why finding shortest paths between caterpillar trees builds the basis for understanding the difference between $\nni$ and $\rnni$.
Further details on our computational results can be found in Section~\ref{section:computation}.

Before we start analysing shortest paths between caterpillar trees, we introduce further notations.
A path $p$ between two caterpillar trees $T$ and $R$ that only consists of caterpillar trees is called \emph{caterpillar path}.
If $p$ is shortest among all caterpillar paths between these trees we call it \emph{shortest caterpillar path}, and the length of $p$ will be denoted by $d_c(T,R)$ and is called \emph{caterpillar distance}.

Let us compare shortest paths between caterpillar trees in $\nni$ with those in $\rnni$.
The essential difference is that building a cherry and moving it may give a path that is shorter than a path where the taxa move separately in $\nni$ as it is illustrated out in Figure~\ref{fig:NNI_vs_RNNI}.
There is only one move needed in $\nni$ to build and resolve a cherry, respectively, while the number of moves for moving the cherry is the same as for moving a single taxon.
However, in $\rnni$ space, there are additional rank moves needed for the internal node of the cherry before an $\nni$ move resolves it, because $\nni$ moves are only allowed on edges of length one in $\rnni$.
Therefore, a path where taxa move separately and no second cherry is built, is shortest in $\rnni$.

@@

The above observation already suggests that there is always a caterpillar path that is shortest path between two arbitrary caterpillar trees in $\rnni$.
Indeed, we are going to prove Theorem~\ref{thm:caterpillar_convex} within this section, which states that the set of caterpillar trees is convex in $\rnni$.
Note that this is not true for the $\nni$ space, which is clear from the example in Figure~\ref{fig:NNI_vs_RNNI}.

One can see caterpillar trees as sequences of taxa, ordered non-decreasing according to the ranks of their parents.
$\nni$ moves between caterpillar trees, which are the only possible $\rnni$ moves between those trees, coincide with exchanges of two neighboured elements in the sequence.
In these sequences the first two taxa are somehow equivalent since they build the unique cherry.
$\nni$ moves that change the cherry of a tree refer to exchanges of the first or the second element with the third element of the sequence while exchanges of the first two elements do not change the corresponding tree.
With this observation we can provide an algorithm for computing shortest caterpillar paths by a simple modification of the commonly known \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi}.
Throughout this paper this algorithm will be called $\csort$.

Following the observations above, only swaps of the first three elements of a sequence with Bubble Sort need to be adapted for the computation of shortest caterpillar paths.
Let us consider the first three elements $a,b,c$ which appear in this order in a sequence representing a caterpillar tree on a caterpillar path between trees $T$ and $R$.
If the rank of $\parent(c)$ is higher than the ranks of $\parent(a)$ and $\parent(b)$ in $R$, there are no swaps needed for these taxa.
If just one of the two taxa $a$ and $b$ is on the other side of $c$ in $R$, this one swaps position with $c$ in the sequence.
But if both $a$ and $b$ are on the other side of $c$ in $R$, the one that has the parent with highest rank of all these three taxa in $R$ swaps with $c$ at first.
This makes sure that there is no additional swap of $a$ and $b$ needed on the path to $R$.
\todo{Do we need to explain Bubble Sort?}

Lemma~\ref{lemma:caterpillar_dist=diameter} regards caterpillar trees with distance $\frac{(n-1)(n-2)}{2}$ and will be needed for proving the convexity of the set of caterpillar trees in $\rnni$.

@@

\begin{proof}
    Let $T$ and $R$ be caterpillar trees.
    Let us first prove that if $d(T,R) = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.
    This easily follows from algorithm $\csort$:
    In each loop $i=1, \ldots, n-2$ of the algorithm there can be at most $n-1-i$ exchanges of taxa.
    It follows that the maximum length of a path computed by this algorithm is $\sum\limits_{k=1}^{n-2} k = \frac{(n-1)(n-2)}{2}$.
    Since there cannot be a caterpillar path shorter than $\frac{(n-1)(n-2)}{2}$ if it is $d(T,R) = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,R) = \Delta(\rnniu)$.

    For proving the other direction of the statement, let now $T$ and $R$ have caterpillar distance $d_c(T,R) =  \frac{(n-1)(n-2)}{2}$.
    We prove $d(T,R) = \frac{(n-1)(n-2)}{2}$ by induction on the number of taxa $n$ of $T$ and $R$:
    Without loss of generality we can assume that $T$ is the caterpillar tree $(\ldots ((1,2),3), \ldots, n)$.
    Let $R$ be a caterpillar tree with distance $\frac{(n-1)(n-2)}{2}$ to $T$.

    Induction basis: $n=3$\\
    All three trees have distance $1$ from another and are caterpillar trees, hence it is $d(T,R) = d_c(T,R)$ for all trees $T,R$ on $n=3$ taxa.

    Induction step: $n \to n+1$\\
    Induction hypothesis: For trees $T, R$ on $n$ taxa it is $d(T, R) = \frac{n(n-1)}{2}$.\\
    Let $T, R$ be trees on $n+1$ taxa with caterpillar distance $\frac{(n-1)(n-2)}{2}$.
    Therefore, each of the $\frac{(n-1)(n-2)}{2}$ pairs of taxa that is considered when running $\csort$ to compute a path $p$ from $R$ to $T$ exchanges.
    From this we can conclude that taxon $n+1$, whose parent has highest rank in $T$, is part of the cherry of $R$.
    Let us consider the trees $T{\big|}_n$ to $R{\big|}_n$ that result from $T$ and $R$ by deleting taxon $n+1$.
    $\csort$ computes a path from $T{\big|}_n$ to $R{\big|}_n$ that contains the same exchanges of taxa as $p$, except for the $n-1$ first ones on $p$ that involve taxon $n+1$.
    Hence, the caterpillar distance is $d_c(T{\big|}_n, R{\big|}_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
    So we can apply the induction hypothesis on $T{\big|}_n$ and $R{\big|}_n$ and know that $d(T{\big|}_n,R{\big|}_n) = \frac{(n-1)(n-2)}{2}$.

    According to Lemma~\ref{lemma:distance_delete_taxon}, the distance between $T{\big|}_n$ and $R{\big|}_n$ is at least $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))|$.
    We also know $|\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))| = n-1$
    Suppose, contrary to our claim, that there is a path from $T$ to $R$ of length less than $\Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2}$.
    By the above, it follows that $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))| < \frac{n(n-1)}{2} - (n-1) < \frac{(n-1)(n-2)}{2}$.
    Since this is a contradiction to $d(T{\big|}_n, R{\big|}_n) = \frac{(n-1)(n-2)}{2}$, it is proven that $d(T,R) = \frac{n(n-1)}{2} $.

\end{proof}

@@

\begin{proof}
    We prove the Theorem by backwards induction on the length $l$ of a shortest caterpillar path between two caterpillar trees $T$ and $R$.

    Induction basis: $l = \frac{(n-1)(n-2)}{2}$\\
    With Lemma~\ref{lemma:caterpillar_dist=diameter} one can see that for each pair of caterpillar trees $T$, $R$ with caterpillar distance $d_c(T,R) = \frac{(n-1)(n-2)}{2}$, it is $d(T,R) = d_c(T,R)$, which means that there is a shortest path between $T$ and $R$ that is a caterpillar path.

    Induction step: $l+1 \to l$\\
    Induction hypothesis: For each pair of caterpillar trees with caterpillar distance $l+1$ there is a shortest path between these trees that only consists of caterpillar trees, which means that the distance between these trees is $l+1$.
    Let $T$ and $R$ be caterpillar trees with caterpillar distance $l$.
    We consider the shortest caterpillar path computed by $\csort$.
    Since $l < \frac{(n-1)(n-2)}{2}$, there is one loop $k$ when running $\csort$ where less than $n-2-k$ pairs taxa swap positions.
    It follows that there is a taxon $a$ that has parent with rank $n-2-k$ in $T$ and does not need to move all the way up starting at the cherry in loop $k$.
    Let $b$ be the taxon whose parent has rank one less than the parent of $a$ in the tree at the beginning of loop number $k$.
    We can assume without loss of generality that loop $k$ is the first loop in which the number of swaps is not maximal.
    It follows that $a$ and $b$ must be neighboured taxa in $R$.
    Let us now consider caterpillar tree $R'$ resulting from $R$ by exchanging taxa $a$ and $b$.
    It is clear that $R'$ has caterpillar distance $d_c(T,R') = d_c(T,R) + 1 = l+1$ to $T$ and moreover, $d(R,R') = 1$.
    Hence we can apply the induction hypothesis on $T$ and $R'$ and see that $d(T,R') = l+1$.
    If it were $d(T,R) < l$, we would have $d(T,R') \leq d(T,R) + d(R,R') < l + 1$ which contradicts the induction hypothesis.
    We can conclude $d(T,R) = l$, which completes the proof.
\end{proof}

>>>>

\begin{lemma}
    $\Delta(\rnni) \leq \frac{(n-1)(n-2)}{2}$, where $\Delta(G) = \max\limits_{u,v \in V(G)}d(u,v)$ denotes the diameter of graph $G$.
    \label{lemma:diameter_bound}
\end{lemma}

@@

In this section we restrict our attention to the set of \emph{caterpillar trees} where each internal node has at least one child that is a leaf.
This type of tree is of particular interest in $\rnni$ because shortest paths between such trees in $\rnni$ differ from those in classical $\nni$ space.
This in combination with the importance of caterpillar trees in the proof of $\np$-hardness of $\nni$ distances in \autocite{Dasgupta2000-xa} suggests that understanding the geometry of paths between caterpillar trees helps investigating the complexity of computing $\rnni$ distances.
In particular, we are going to prove that one can always find a shortest path between two caterpillar trees that contains caterpillar trees only, meaning that the set of caterpillar trees is convex in $\rnni$.
We discovered this by computing caterpillar paths using an algorithm to compute shortest paths within the set of caterpillar trees and comparing them with shortest path in $\rnni$.
This algorithm will be explained in detail later in this section, after demonstrating why finding shortest paths between caterpillar trees builds the basis for understanding the difference between $\nni$ and $\rnni$.
Further details on our computational results can be found in Section~\ref{section:computation}.

Before we start analysing shortest paths between caterpillar trees, we need to introduce further notations.
A path $p$ between two caterpillar trees $T$ and $R$ that only consists of caterpillar trees is called \emph{caterpillar path}.
If $p$ is shortest among all caterpillar paths between these trees we call it \emph{shortest caterpillar path}, and the length of $p$ will be denoted by $d_c(T,R)$ and called \emph{caterpillar distance}.

We now compare shortest paths between caterpillar trees in $\nni$ with those in $\rnni$.
In $\nni$, building a cherry and moving it may give a path that is shorter than a path where the taxa move separately in as it is illustrated out in Figure~\ref{fig:NNI_vs_RNNI}.
There is only one move needed to build and resolve a cherry, respectively, while the number of moves for moving the cherry is the same as for moving a single taxon in $\nni$.
But in $\rnni$ space there are additional rank moves needed to change the rank of the internal node of the cherry before an $\nni$ move resolves it.
This is due to the fact that $\nni$ moves are only allowed on edges of length one in $\rnni$.
Therefore, a path where taxa move separately and no second cherry is built, is shortest in $\rnni$.

@@

The above observation suggests that there is always a caterpillar path that is shortest path between two arbitrary caterpillar trees in $\rnni$.
Indeed, we are going to prove Theorem~\ref{thm:caterpillar_convex} within this section, which states that the set of caterpillar trees is convex in $\rnni$.
Note that this is not true for the $\nni$ space, which is clear from the example in Figure~\ref{fig:NNI_vs_RNNI}.

We next introduce a unique representation of caterpillar trees to show how one can use the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi} to compute caterpillar paths.
One can see caterpillar trees as sequences of taxa, ordered non-decreasing according to the ranks of their parents.
$\nni$ moves between caterpillar trees, which are the only possible $\rnni$ moves between those trees, coincide with exchanges of two neighboured elements in the sequence.
In these sequences the first two taxa are somehow equivalent since they build the unique cherry which means they share their parent with rank one.
Therefore, we need to consider $\nni$ moves that change the cherry of a tree separately.
These moves refer to exchanges of the first or the second element with the third element of the sequence, while exchanges of the first two elements do not change the corresponding tree.
With this observation we can provide an algorithm for computing shortest caterpillar paths by a simple modification of the commonly known Bubble Sort algorithm.
Throughout this paper this algorithm will be called $\csort$.

Following the observations above, only swaps of the first three elements of a sequence with Bubble Sort need to be adapted for the computation of shortest caterpillar paths.
Let us consider the first three elements $a,b,c$, appearing in this order in a sequence representing a caterpillar tree on a caterpillar path between trees $T$ and $R$.
If the rank of $\parent(c)$ is higher than the ranks of $\parent(a)$ and $\parent(b)$ in $R$, there are no swaps needed for these taxa.
If just one of the two taxa $a$ and $b$ is on the other side of $c$ in $R$, this taxon swaps position with $c$ in the sequence.
And if both $a$ and $b$ are on the other side of $c$ in $R$, the taxon that has the parent with highest rank of all these three taxa in $R$ swaps with $c$ at first.
This makes sure that there is no additional swap of $a$ and $b$ needed on the subsequent path to $R$.
\todo{Do we need to explain Bubble Sort?}

Lemma~\ref{lemma:caterpillar_dist=diameter} regards caterpillar trees with distance $\frac{(n-1)(n-2)}{2}$ and will be needed for proving the convexity of the set of caterpillar trees in $\rnni$.
Notice that we by Lemma~\ref{lemma:diameter_bound} $\frac{(n-1)(n-2)}{2}$ is the upper bound for the diameter of $\rnni$.

@@

\begin{proof}
    Let $T$ and $R$ be caterpillar trees.
    Let us first prove that if $d(T,R) = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.
    This easily follows from algorithm $\csort$:
    In each loop $i=1, \ldots, n-2$ of the algorithm there can be at most $n-1-i$ exchanges of taxa.
    It follows that the maximum length of a path computed by this algorithm is $\sum\limits_{k=1}^{n-2} k = \frac{(n-1)(n-2)}{2}$.
    Since there cannot be a caterpillar path shorter than $\frac{(n-1)(n-2)}{2}$ if it is $d(T,R) = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,R) =  \frac{(n-1)(n-2)}{2}$.

    For proving the other direction of the statement, let now $T$ and $R$ have caterpillar distance $d_c(T,R) =  \frac{(n-1)(n-2)}{2}$.
    We prove $d(T,R) = \frac{(n-1)(n-2)}{2}$ by induction on the number of taxa $n$ of $T$ and $R$:
    Without loss of generality we can assume that $T$ is the caterpillar tree $(\ldots ((1,2),3), \ldots, n)$ or $(\ldots ((1,2),3), \ldots, n+1)$ in the induction step.
    Let $R$ be a caterpillar tree with distance $\frac{(n-1)(n-2)}{2}$ to $T$.

    Induction basis: $n=3$\\
    There are three trees on $3$ taxa that have distance $1$ from another and are caterpillar trees.
    Hence it is $d(T,R) = d_c(T,R) = 1$ for all pairs of trees $T,R$ on $n=3$ taxa.

    Induction step: $n \to n+1$\\
    Induction hypothesis: For trees $T, R$ on $n$ taxa it is $d(T, R) = \frac{(n-1)(n-2)}{2}$.\\
    Let $T, R$ be trees on $n+1$ taxa with caterpillar distance $\frac{n(n-1)}{2}$.
    Therefore, each of the $\frac{n(n-1)}{2}$ pairs of taxa that is considered when running $\csort$ to compute a path $p$ from $R$ to $T$ exchanges.
    From this we can conclude that taxon $n+1$, whose parent has highest rank in $T$, is part of the cherry of $R$. \todo{Do we need to explain this further (figure)?}
    Let us consider the trees $T{\big|}_n$ to $R{\big|}_n$ that result from $T$ and $R$ by deleting taxon $n+1$.
    $\csort$ computes a path from $T{\big|}_n$ to $R{\big|}_n$ that contains the same exchanges of taxa as $p$, except for the $n-1$ ones on $p$ that involve taxon $n+1$.
    Hence, the caterpillar distance is $d_c(T{\big|}_n, R{\big|}_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
    So we can apply the induction hypothesis on $T{\big|}_n$ and $R{\big|}_n$ and know that $d(T{\big|}_n,R{\big|}_n) = \frac{(n-1)(n-2)}{2}$.

    According to Lemma~\ref{lemma:distance_delete_taxon}, the distance between $T{\big|}_n$ and $R{\big|}_n$ is at least $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))|$.
    We also know $|\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))| = n-1$.
    Suppose, contrary to our claim, that there is a path from $T$ to $R$ of length less than $\frac{n(n-1)}{2}$.
    By the above, it follows that $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))| < \frac{n(n-1)}{2} - (n-1) < \frac{(n-1)(n-2)}{2}$.
    Since this is a contradiction to $d(T{\big|}_n, R{\big|}_n) = \frac{(n-1)(n-2)}{2}$, it is proven that $d(T,R) = \frac{n(n-1)}{2} $.
\end{proof}

@@

\begin{proof}
    We prove this Theorem by backwards induction on the length $l$ of a shortest caterpillar path between two caterpillar trees $T$ and $R$.
    Since we use backwards induction, we need to consider the maximum distance between any two caterpillar trees as base case, which is $\frac{(n-1)(n-2)}{2}$ as we saw above.

    Induction basis: $l = \frac{(n-1)(n-2)}{2}$\\
    With Lemma~\ref{lemma:caterpillar_dist=diameter} one can see that for each pair of caterpillar trees $T$, $R$ with caterpillar distance $d_c(T,R) = \frac{(n-1)(n-2)}{2}$, it is $d(T,R) = d_c(T,R)$.
    Therefore, there is a shortest path between $T$ and $R$ that is a caterpillar path.

    Induction step: $l+1 \to l$\\
    Induction hypothesis: For each pair of caterpillar trees with caterpillar distance $l+1$ there is a shortest path between these trees that only consists of caterpillar trees, which means that the distance between these trees is $l+1$.
    Let $T$ and $R$ be caterpillar trees with caterpillar distance $l$.
    We consider the shortest caterpillar path computed by $\csort$.
    Since $l < \frac{(n-1)(n-2)}{2}$, there is one loop $k$ when running $\csort$ where less than $n-2-k$ taxon pairs swap positions.
    We assume that this loop number $k$ is the first one in which this occurs.
    This situation is illustrated in the top of Figure~\ref{fig:cat_convex_seq_example}.
    Whenever the number of swaps of taxa or, in terms of sequences, neighboured elements in the sequence is maximum in a loop, one of the two first elements swaps position with $n-2-k$ subsequent elements of the sequence.
    Since we assume that the number of swaps in loop $k$ is not maximal, this does not happen.
    Let us assume that $a$ is the first element that is compared with its left neighbour in loop $k$ but does not swap with it.
    Then the left neighbour $b$ of $a$ in $T$ is below $a$ in $R$ as well.
    Therefore, the tree $T'$ which can be received from $T$ by exchanging $a$ and $b$ must have caterpillar distance $d_c(T',R) = d_c(T,R) + 1 = l + 1$ to $R$.
    For an illustration of the $k^{th}$ loop when performing $\csort$ between $T$ and $R$ or $T'$ and $R$, respectively, consider Figure~\ref{fig:cat_convex_seq_example}.

    \begin{figure}[H]
    	\centering
    	\includegraphics[width=\textwidth]{cat_convex_seq_example}
    	\caption{Loop number $k$ of $\csort$ between $T$ and $R$ as described in proof of Theorem~\ref{thm:caterpillar_convex}.
        The trees are represented as sequences that are being ordered by $\csort$.
        Right before loop number $k$, $x_1, \ldots, x_{k-1}$ are already at the same position as in $R$.
        At the top: sequences between $T$ and $R$ -- $x_k$ and $b$ swap in loop number $k$, but $x_k$ and $a$ not.
        At the bottom: sequences between $T'$ and $R$ -- $a$ and $b$ swap in loop number $k$; $x_k$ and $b$ will swap later}
    	\label{fig:cat_convex_seq_example}
    \end{figure}

    Since it is clear that $T'$ has caterpillar distance $d_c(T',R) = l+1$ to $T$, we can apply the induction hypothesis on $T'$ and $R$ and see that $d(T',R) = l+1$.
    Furthermore, it is $d(T,T') = 1$.
    If it now were $d(T,R) < l$, it would hold $d(T',R) \leq d(T,R) + d(T,T') < l + 1$ which contradicts the induction hypothesis.
    We can conclude $d(T,R) = l$, which completes the proof.
\end{proof}
